Stochastic Variance-Reduced Policy Gradient

### 2018-05-30 svrpg.tex
Updated

\vspace{-0.05in}
\section{Introduction}
\vspace{-0.05in}
%Similarities between RL and SL
On a very general level, artificial intelligence addresses the problem of an agent that must select the right actions to solve a task. The approach of Reinforcement Learning~\citep{sutton1998reinforcement} is to learn the best actions by direct interaction with the environment and evaluation of the performance in the form of a reward signal. This makes RL fundamentally different from Supervised Learning (SL), where correct actions are explicitly prescribed by a human teacher (\eg for classification, in the form of class labels). However, the two approaches share many challenges and tools. The problem of estimating a model from samples, which is at the core of SL, is equally fundamental in RL, whether we choose to model the environment, a value function, or directly a policy defining the agent's behaviour. Furthermore, when the tasks are characterized by large or continuous state-action spaces, RL needs the powerful function approximators (\eg neural networks) that are the main subject of study of SL.
% SL needs SVRG so RL
In a typical SL setting, a performance function $J(\vtheta)$ has to be optimized \wrt to model parameters $\vtheta$. The set of data that are available for training is often a subset of all the cases of interest, which may even be infinite, leading to optimization of finite sums that approximate the expected performance over an unknown data distribution. When generalization to the complete dataset is not taken into consideration, we talk about Empirical Risk Minimization (ERM). Even in this case, stochastic optimization is often used for reasons of efficiency. The idea of stochastic gradient (SG) ascent \cite{nesterov2013introductory} is to iteratively focus on a random subset of the available data to obtain an approximate improvement direction. At the level of the single iteration, this can be much less expensive than taking into account all the data. However, the sub-sampling of data is a source of variance that can potentially compromise convergence, so that per-iteration efficiency and convergence rate must be traded off with proper handling of meta-parameters.
Variance-reduced gradient algorithms such as SAG \cite{roux2012stochastic}, SVRG \cite{johnson2013accelerating} and SAGA \cite{defazio2014saga} offer better ways of solving this trade-off, with significant results both in theory and practice. Although designed explicitly for ERM, these algorithms address a problem that affects more general machine learning problems.

In RL, stochastic optimization is rarely a matter of choice, since data must be actively sampled by interacting with an initially unknown environment. In this scenario, limiting the variance of the estimates is a necessity that cannot be avoided, which makes variance-reduced algorithms very interesting.
% Potential advantages of SVRG
Among RL approaches, policy gradient \cite{sutton2000policy} is the one that bears the closest similarity to SL solutions. The fundamental principle of these methods is to optimize a parametric policy through stochastic gradient ascent. Compared to other applications of SG, the cost of collecting samples can be very high since it requires to interact with the environment. This makes SVRG-like methods potentially much more efficient than, \eg batch learning.
% Difficulties in RL \wrt SL
Unfortunately, RL has a series of difficulties that are not present in ERM. First, in SL the objective can often be designed to be strongly concave (we aim to maximize). This is not the case for RL, so we have to deal with non-concave objective functions. Then, as mentioned before, the dataset is not initially available and may even be infinite, which makes approximations unavoidable. This rules out SAG and SAGA because of their storage requirements, which leaves SVRG as the most promising choice. Finally, the distribution used to sample data is not under direct control of the algorithm designer, but it is a function of policy parameters that changes over time as the policy is optimized, which is a form of non-stationarity.
% Related work?
SVRG has been used in RL as an efficient technique for optimizing the per-iteration problem in Trust-Region Policy Optimization~\citep{xu2017svrgtrpo} or for policy evaluation~\citep{du2017svrgpe}.
In both the cases, the optimization problems faced resemble the SL scenario and are not affected by all the previously mentioned issues.

%Paper structure
After providing background on policy gradient and SVRG in Section \ref{sec:pre}, in Section \ref{sec:alg} we propose SVRPG, a variant of SVRG for the policy gradient framework, addressing all the difficulties mentioned above. In Section \ref{sec:conv} we provide convergence guarantees for our algorithm, and we show a convergence rate that has an $O(\nicefrac{1}{T})$ dependence on the number $T$ of iterations. In Section \ref{sec:stopping} we suggest how to set the meta-parameters of SVRPG, while in Section \ref{sec:prac} we discuss some practical variants of the algorithm. Finally, in Section \ref{sec:exp} we empirically evaluate the performance of our method on popular continuous RL tasks.

@@

\vspace{-0.05in}
\subsection{Policy Gradient}\label{subsec:PolicyGradient}
\vspace{-0.05in}
A Reinforcement Learning task~\citep{sutton1998reinforcement} can be modelled with a discrete-time continuous Markov Decision Process (MDP) $M = \{\Sspace,\Aspace,\Transition,\Reward,\gamma,\rho\}$, where $\Sspace$ is a continuous state space; $\Aspace$ is a continuous action space; $\Transition$ is a Markovian transition model, where $\Transition(s'|s,a)$ defines the transition density from state $s$ to $s'$ under action $a$; $\Reward$ is the reward function, where $\Reward(s,a) \in [-R,R]$ is the expected reward for state-action pair $(s,a)$;
% and $R$ is the maximum absolute-value reward;
$\gamma\in[0,1)$ is the discount factor; and $\rho$ is the initial state distribution.
The agent's behaviour is modelled as a policy $\pi$, where $\pi(\cdot|s)$ is the density distribution over $\Aspace$ in state $s$.
% We consider episodic tasks, \ie tasks composed of episodes of length $H$, also called time horizon.
% In this setting, a trajectory $\tau$ is the sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following some stationary policy in a given episode, where $s_0$ is always sampled from the initial state distribution $\rho$.
We consider episodic MDPs with effective horizon $H$.\footnote{The episode duration is a random variable, but the optimal policy can reach the target state (\ie absorbing state) in at most $H$ steps. This has not to be confused with a finite horizon problem where the optimal policy is non-stationary.} In this setting, we can limit our attention to trajectories of length $H$. A trajectory $\tau$ is a sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following a stationary policy, where $s_0 \sim \rho$.
We denote with $p(\tau|\pi)$ the density distribution induced by policy $\pi$ on the set $\Tspace$ of all possible trajectories (see Appendix~\ref{A:gradient_estimators} for the definition), and with $\Reward(\tau)$ the total discounted reward provided by trajectory $\tau$:
%
$\Reward(\tau) = \sum_{t=0}^{H-1}\gamma^t\Reward(s_t,a_t).$
%
Policies can be ranked based on their expected total reward: $J(\pi) = \EVV[\tau \sim p(\cdot|\pi)]{\Reward(\tau)|M}$.
Solving an MDP $M$ means finding $\pi^* \in \argmax_{\pi} \{J(\pi)\}$.

@@

\vspace{-0.05in}
\subsection{Stochastic Variance-Reduced Gradient}\label{sec:svrg}
\vspace{-0.05in}
Finite-sum optimization is the problem of maximizing an objective function $f(\vtheta)$ which can be decomposed into the sum or average of a finite number of functions $z_i(\cdot|\vtheta)$:
\begin{align*}
        \max_{\vtheta} \left\{ f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}z_i(\vtheta)\right\}.
\end{align*}
This kind of optimization is very common in machine learning, where each $z_i$ may correspond to a data sample $x_i$ from a dataset $\mathcal{D}_N$ of size $N$ (\ie $z_i(\vtheta) = z(x_i|\vtheta)$). 
%% In this case, we adopt the following, more meaningful notation:
%% \begin{align*}
%% \max_{\vtheta} f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}g(x_i \vert \vtheta).
%% \end{align*}
A common requirement is that $z$ must be smooth and concave in $\vtheta$.\footnote{Note that we are considering a maximization problem instead of the classical minimization one.} 
Under this hypothesis, full gradient (FG) ascent~\citep{cauchy1847methode} with a constant step size achieves a linear convergence rate in the number $T$ of iterations (\ie parameter updates)~\citep{nesterov2013introductory}.
However, each iteration requires $N$ gradient computations, which can be too expensive for large values of $N$. Stochastic Gradient (SG) ascent~\citep[\eg][]{robbins1951stochastic,bottou2004large} overcomes this problem by sampling a single sample $x_i$ per iteration, but a vanishing step size is required to control the variance introduced by sampling. As a consequence, the lower per-iteration cost is paid with a worse, sub-linear convergence rate \cite{nemirovskii1983problem}.
Starting from SAG, a series of variations to SG have been proposed to achieve a better trade-off between convergence speed and cost per iteration: \eg SAG~\citep{roux2012stochastic}, SVRG~\cite{johnson2013accelerating}, SAGA~\cite{defazio2014saga}, Finito~\cite{defazio2014finito}, and MISO~\cite{mairal2015incremental}. 
The common idea is to reuse past gradient computations to reduce the variance of the current estimate.
In particular, Stochastic Variance-Reduced Gradient (SVRG) is often preferred to other similar methods for its limited storage requirements, which is a significant advantage when deep and/or wide neural networks are employed.

The idea of SVRG (Algorithm~\ref{alg:svrg}) is to alternate full and stochastic gradient updates. 
Each $m = \mathcal{O}(N)$ iterations, a snapshot $\widetilde{\vtheta}$ of the current parameter is saved together with its full gradient $\nabla f(\widetilde{\vtheta}) = \frac{1}{N} \sum_i \nabla z(x_i|\widetilde{\vtheta})$.
Between snapshots, the parameter is updated with $\blacktriangledown f(\vtheta)$, a gradient estimate corrected using stochastic gradient. For any $t \in \{0,\ldots,m-1\}$:
\begin{equation}\label{E:svrg.gradient.correction}
        \blacktriangledown f(\vtheta_{t}) := v_t = \nabla f(\wt{\vtheta}) + \nabla z(x | \vtheta_t) - \nabla z(x | \wt{\vtheta}),
\end{equation} 
where $x$ is sampled uniformly at random from $\mathcal{D}_N$ (\ie $x \sim \mathcal{U}(\mathcal{D}_N)$).
Note that $t=0$ corresponds to a FG step (\ie $\blacktriangledown f(\vtheta_0) = \nabla f(\wt{\vtheta})$) since $\vtheta_0 := \wt{\vtheta}$.
The corrected gradient $\blacktriangledown f(\vtheta)$ is an unbiased estimate of $\nabla f(\vtheta)$, and it is able to control the variance introduced by sampling even with a fixed step size, achieving a linear convergence rate without resorting to a plain full gradient.

More recently, some extensions of variance reduction algorithms to the non-concave objectives have been proposed~\citep[\eg][]{allen2016variance,reddi2016stochastic,reddi2016fast}. In this scenario, $f$ is typically required to be L-smooth, \ie $\norm[2]{\nabla f(\vtheta') - \nabla f(\vtheta)} \leq L\norm[2]{\vtheta'-\vtheta}$ for each $\vtheta,\vtheta'\in\realspace^n$ and for some Lipschitz constant $L$. Under this hypothesis, the convergence rate of SG is $O(\nicefrac{1}{\sqrt{T}})$ \cite{ghadimi2013stochastic}, \ie $T=O(\nicefrac{1}{\epsilon^2})$ iterations are required to get $\norm[2]{\nabla f(\vtheta)}^2\leq\epsilon$. Again, SVRG achieves the same rate as FG \cite{reddi2016stochastic}, which is $O(\frac{1}{T})$ in this case \cite{nesterov2013introductory}. The only additional requirement is to select $\vtheta^*$ uniformly at random among all the $\vtheta_k$ instead of simply setting it to the final value.

>>>>

\vspace{-0.05in}
\section{Introduction}
\vspace{-0.05in}
%Similarities between RL and SL
On a very general level, artificial intelligence addresses the problem of an agent that must select the right actions to solve a task. The approach of Reinforcement Learning (RL)~\citep{sutton1998reinforcement} is to learn the best actions by direct interaction with the environment and evaluation of the performance in the form of a reward signal. This makes RL fundamentally different from Supervised Learning (SL), where correct actions are explicitly prescribed by a human teacher (\eg for classification, in the form of class labels). However, the two approaches share many challenges and tools. The problem of estimating a model from samples, which is at the core of SL, is equally fundamental in RL, whether we choose to model the environment, a value function, or directly a policy defining the agent's behaviour. Furthermore, when the tasks are characterized by large or continuous state-action spaces, RL needs the powerful function approximators (\eg neural networks) that are the main subject of study of SL.
% SL needs SVRG so RL
In a typical SL setting, a performance function $J(\vtheta)$ has to be optimized \wrt to model parameters $\vtheta$. The set of data that are available for training is often a subset of all the cases of interest, which may even be infinite, leading to optimization of finite sums that approximate the expected performance over an unknown data distribution. When generalization to the complete dataset is not taken into consideration, we talk about Empirical Risk Minimization (ERM). Even in this case, stochastic optimization is often used for reasons of efficiency. The idea of stochastic gradient (SG) ascent \cite{nesterov2013introductory} is to iteratively focus on a random subset of the available data to obtain an approximate improvement direction. At the level of the single iteration, this can be much less expensive than taking into account all the data. However, the sub-sampling of data is a source of variance that can potentially compromise convergence, so that per-iteration efficiency and convergence rate must be traded off with proper handling of meta-parameters.
Variance-reduced gradient algorithms such as SAG \cite{roux2012stochastic}, SVRG \cite{johnson2013accelerating} and SAGA \cite{defazio2014saga} offer better ways of solving this trade-off, with significant results both in theory and practice. Although designed explicitly for ERM, these algorithms address a problem that affects more general machine learning problems.

In RL, stochastic optimization is rarely a matter of choice, since data must be actively sampled by interacting with an initially unknown environment. In this scenario, limiting the variance of the estimates is a necessity that cannot be avoided, which makes variance-reduced algorithms very interesting.
% Potential advantages of SVRG
Among RL approaches, policy gradient \cite{sutton2000policy} is the one that bears the closest similarity to SL solutions. The fundamental principle of these methods is to optimize a parametric policy through stochastic gradient ascent. Compared to other applications of SG, the cost of collecting samples can be very high since it requires to interact with the environment. This makes SVRG-like methods potentially much more efficient than, \eg batch learning.
% Difficulties in RL \wrt SL
Unfortunately, RL has a series of difficulties that are not present in ERM. First, in SL the objective can often be designed to be strongly concave (we aim to maximize). This is not the case for RL, so we have to deal with non-concave objective functions. Then, as mentioned before, the dataset is not initially available and may even be infinite, which makes approximations unavoidable. This rules out SAG and SAGA because of their storage requirements, which leaves SVRG as the most promising choice. Finally, the distribution used to sample data is not under direct control of the algorithm designer, but it is a function of policy parameters that change over time as the policy is optimized, which is a form of non-stationarity.
% Related work?
SVRG has been used in RL as an efficient technique for optimizing the per-iteration problem in Trust-Region Policy Optimization~\citep{xu2017svrgtrpo} or for policy evaluation~\citep{du2017svrgpe}.
In both the cases, the optimization problems faced resemble the SL scenario and are not affected by all the previously mentioned issues.

%Paper structure
After providing background on policy gradient and SVRG in Section \ref{sec:pre}, we propose SVRPG, a variant of SVRG for the policy gradient framework, addressing all the difficulties mentioned above (see Section \ref{sec:alg}). In Section \ref{sec:conv} we provide convergence guarantees for our algorithm, and we show a convergence rate that has an $O(\nicefrac{1}{T})$ dependence on the number $T$ of iterations. In Section \ref{sec:stopping} we suggest how to set the meta-parameters of SVRPG, while in Section \ref{sec:prac} we discuss some practical variants of the algorithm. Finally, in Section \ref{sec:exp} we empirically evaluate the performance of our method on popular continuous RL tasks.

@@

\vspace{-0.05in}
\subsection{Policy Gradient}\label{subsec:PolicyGradient}
\vspace{-0.05in}
A Reinforcement Learning task~\citep{sutton1998reinforcement} can be modelled with a discrete-time continuous Markov Decision Process (MDP) $M = \{\Sspace,\Aspace,\Transition,\Reward,\gamma,\rho\}$, where $\Sspace$ is a continuous state space; $\Aspace$ is a continuous action space; $\Transition$ is a Markovian transition model, where $\Transition(s'|s,a)$ defines the transition density from state $s$ to $s'$ under action $a$; $\Reward$ is the reward function, where $\Reward(s,a) \in [-R,R]$ is the expected reward for state-action pair $(s,a)$;
% and $R$ is the maximum absolute-value reward;
$\gamma\in[0,1)$ is the discount factor; and $\rho$ is the initial state distribution.
The agent's behaviour is modelled as a policy $\pi$, where $\pi(\cdot|s)$ is the density distribution over $\Aspace$ in state $s$.
% We consider episodic tasks, \ie tasks composed of episodes of length $H$, also called time horizon.
% In this setting, a trajectory $\tau$ is the sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following some stationary policy in a given episode, where $s_0$ is always sampled from the initial state distribution $\rho$.
We consider episodic MDPs with effective horizon $H$.\footnote{The episode duration is a random variable, but the optimal policy can reach the target state (\ie absorbing state) in less than $H$ steps. This has not to be confused with a finite horizon problem where the optimal policy is non-stationary.} In this setting, we can limit our attention to trajectories of length $H$. A trajectory $\tau$ is a sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following a stationary policy, where $s_0 \sim \rho$.
We denote with $p(\tau|\pi)$ the density distribution induced by policy $\pi$ on the set $\Tspace$ of all possible trajectories (see Appendix~\ref{A:gradient_estimators} for the definition), and with $\Reward(\tau)$ the total discounted reward provided by trajectory $\tau$:
%
$\Reward(\tau) = \sum_{t=0}^{H-1}\gamma^t\Reward(s_t,a_t).$
%
Policies can be ranked based on their expected total reward: $J(\pi) = \EVV[\tau \sim p(\cdot|\pi)]{\Reward(\tau)|M}$.
Solving an MDP $M$ means finding $\pi^* \in \argmax_{\pi} \{J(\pi)\}$.

@@

\vspace{-0.05in}
\subsection{Stochastic Variance-Reduced Gradient}\label{sec:svrg}
\vspace{-0.05in}
Finite-sum optimization is the problem of maximizing an objective function $f(\vtheta)$ which can be decomposed into the sum or average of a finite number of functions $z_i(\vtheta)$:
\begin{align*}
        \max_{\vtheta} \left\{ f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}z_i(\vtheta)\right\}.
\end{align*}
This kind of optimization is very common in machine learning, where each $z_i$ may correspond to a data sample $x_i$ from a dataset $\mathcal{D}_N$ of size $N$ (\ie $z_i(\vtheta) = z(x_i|\vtheta)$). 
%% In this case, we adopt the following, more meaningful notation:
%% \begin{align*}
%% \max_{\vtheta} f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}g(x_i \vert \vtheta).
%% \end{align*}
A common requirement is that $z$ must be smooth and concave in $\vtheta$.\footnote{Note that we are considering a maximization problem instead of the classical minimization one.} 
Under this hypothesis, full gradient (FG) ascent~\citep{cauchy1847methode} with a constant step size achieves a linear convergence rate in the number $T$ of iterations (\ie parameter updates)~\citep{nesterov2013introductory}.
However, each iteration requires $N$ gradient computations, which can be too expensive for large values of $N$. Stochastic Gradient (SG) ascent~\citep[\eg][]{robbins1951stochastic,bottou2004large} overcomes this problem by sampling a single sample $x_i$ per iteration, but a vanishing step size is required to control the variance introduced by sampling. As a consequence, the lower per-iteration cost is paid with a worse, sub-linear convergence rate \cite{nemirovskii1983problem}.
Starting from SAG, a series of variations to SG have been proposed to achieve a better trade-off between convergence speed and cost per iteration: \eg SAG~\citep{roux2012stochastic}, SVRG~\cite{johnson2013accelerating}, SAGA~\cite{defazio2014saga}, Finito~\cite{defazio2014finito}, and MISO~\cite{mairal2015incremental}. 
The common idea is to reuse past gradient computations to reduce the variance of the current estimate.
In particular, Stochastic Variance-Reduced Gradient (SVRG) is often preferred to other similar methods for its limited storage requirements, which is a significant advantage when deep and/or wide neural networks are employed.

The idea of SVRG (Algorithm~\ref{alg:svrg}) is to alternate full and stochastic gradient updates. 
Each $m = O(N)$ iterations, a snapshot $\widetilde{\vtheta}$ of the current parameter is saved together with its full gradient $\nabla f(\widetilde{\vtheta}) = \frac{1}{N} \sum_i \nabla z(x_i|\widetilde{\vtheta})$.
Between snapshots, the parameter is updated with $\blacktriangledown f(\vtheta)$, a gradient estimate corrected using stochastic gradient. For any $t \in \{0,\ldots,m-1\}$:
\begin{equation}\label{E:svrg.gradient.correction}
        \blacktriangledown f(\vtheta_{t}) := v_t = \nabla f(\wt{\vtheta}) + \nabla z(x | \vtheta_t) - \nabla z(x | \wt{\vtheta}),
\end{equation} 
where $x$ is sampled uniformly at random from $\mathcal{D}_N$ (\ie $x \sim \mathcal{U}(\mathcal{D}_N)$).
Note that $t=0$ corresponds to a FG step (\ie $\blacktriangledown f(\vtheta_0) = \nabla f(\wt{\vtheta})$) since $\vtheta_0 := \wt{\vtheta}$.
The corrected gradient $\blacktriangledown f(\vtheta)$ is an unbiased estimate of $\nabla f(\vtheta)$, and it is able to control the variance introduced by sampling even with a fixed step size, achieving a linear convergence rate without resorting to a plain full gradient.

More recently, some extensions of variance reduction algorithms to the non-concave objectives have been proposed~\citep[\eg][]{allen2016variance,reddi2016stochastic,reddi2016fast}. In this scenario, $f$ is typically required to be $L$-smooth, \ie $\norm[2]{\nabla f(\vtheta') - \nabla f(\vtheta)} \leq L\norm[2]{\vtheta'-\vtheta}$ for each $\vtheta,\vtheta'\in\realspace^n$ and for some Lipschitz constant $L$. Under this hypothesis, the convergence rate of SG is $O(\nicefrac{1}{\sqrt{T}})$ \cite{ghadimi2013stochastic}, \ie $T=O(\nicefrac{1}{\epsilon^2})$ iterations are required to get $\norm[2]{\nabla f(\vtheta)}^2\leq\epsilon$. Again, SVRG achieves the same rate as FG \cite{reddi2016stochastic}, which is $O(\frac{1}{T})$ in this case \cite{nesterov2013introductory}. The only additional requirement is to select $\vtheta^*$ uniformly at random among all the $\vtheta_k$ instead of simply setting it to the final value ($k$ being the iterations).
